---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
    
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE since API calls require credentials
)
```


`rsynthbio` is an R package that provides a convenient interface to the [Synthesize Bio](https://www.synthesize.bio/) API, allowing users to generate realistic gene expression data based on specified biological conditions. 
This package enables researchers to easily access AI-generated transcriptomic data for various modalities including bulk RNA-seq and single-cell RNA-seq.

Alternatively, you can AI generate datasets from our [web platform](https://app.synthesize.bio/datasets/).

## How to install

You can install `rsynthbio` from CRAN:

```{r installation, eval=FALSE}
install.packages("rsynthbio")
```

If you want the development version, you can install using the `remotes` package to install from GitHub:

```{r github-installation, eval=FALSE}
if (!("remotes" %in% installed.packages())) {
  install.packages("remotes")
}
remotes::install_github("synthesizebio/rsynthbio")
```

Once installed, load the package:

```{r}
library(rsynthbio)
```

## Authentication

Before using the Synthesize Bio API, you need to set up your API token. The package provides a secure way to handle authentication:

```{r auth-secure, eval=FALSE}
# Securely prompt for and store your API token
# The token will not be visible in the console
set_synthesize_token()

# You can also store the token in your system keyring for persistence
# across R sessions (requires the 'keyring' package)
set_synthesize_token(use_keyring = TRUE)
```

Loading your API key for a session. 

```{r, eval=FALSE}
# In future sessions, load the stored token
load_synthesize_token_from_keyring()

# Check if a token is already set
has_synthesize_token()
```

You can obtain an API token by registering at [Synthesize Bio](https://app.synthesize.bio).

### Security Best Practices

For security reasons, remember to clear your token when you're done:

```{r clear-token, eval = FALSE}
# Clear token from current session
clear_synthesize_token()

# Clear token from both session and keyring
clear_synthesize_token(remove_from_keyring = TRUE)
```

Never hard-code your token in scripts that will be shared or committed to version control.

## Basic Usage

### Creating a Query

The first step to generating AI-generated gene expression data is to create a query. The package provides a sample query that you can modify:

```{r query}
# Get a sample query
query <- get_valid_query()

# Inspect the query structure
str(query)
```

The query consists of:

1. `output_modality`: The type of gene expression data to generate (see `get_valid_modalities`)
2. `mode`: The prediction mode (e.g., "mean estimation" or "sample generation")
3. `inputs`: A list of biological conditions to generate data for

We train our models with diverse multi-omics datasets. There are two model modes available today:

+ Mean estimation: These models create a distribution capturing the biological heterogeneity consistent with the supplied metadata. This distribution is then sampled to predict a gene expression distribution that captures measurement error. The mean of that distribution serves as the prediction
+ Sample generation: This model works identically to the mean estimation approach except that the final gene expression distribution is also sampled to generate realistic looking synthetic data that captures error associated with measurement

```{r predict, eval=FALSE}
result <- predict_query(query)
```

This result will be a list of two dataframes: `metadata` and `expression`

### Modifying a Query

You can customize the query to fit your specific research needs:

```{r modify-query}

# Adjust number of samples
query$inputs[[1]]$num_samples <- 10

# Add a new condition
query$inputs[[3]] <- list(
  metadata = list(
    sex = "male",
    sample_type = "primary tissue"
  ),
  num_samples = 3
)
```

The input metadata is a list of lists.

Here are the available metadata fields:

__Biological:__  

- ``age_years``
- ``cell_line_ontology_id``
- ``cell_type_ontology_id``
- ``developmental_stage``
- ``disease_ontology_id``
- ``ethnicity``
- ``genotype``
- ``race``
- ``sample_type`` ("cell line", "organoid", "other", "primary cells", "primary tissue", "xenograft")
- ``sex`` ("male", "female")
- ``tissue_ontology_id``

__Perturbational:__  

- ``perturbation_dose``
- ``perturbation_ontology_id``
- ``perturbation_time``
- ``perturbation_type`` ("coculture","compound","control","crispr","genetic","infection","other","overexpression","peptide or biologic","shrna","sirna")

__Technical:__  

- ``study`` (Bioproject ID)
- ``library_selection`` (e.g., "cDNA", "polyA", "Oligo-dT" - see https://ena-docs.readthedocs.io/en/latest/submit/reads/webin-cli.html#permitted-values-for-library-selection)
- ``library_layout`` ("PAIRED", "SINGLE")
- ``platform`` ("illumina")

### Acceptable Metadata Values

The following are the valid values or expected formats for selected metadata keys:

| Metadata Field             | Requirement / Example                                                                                                                                                                                                 |
|----------------------------|-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `cell_line_ontology_id`    | Requires a [Cellosaurus ID](https://www.cellosaurus.org/).                                                                                                                                                           |
| `cell_type_ontology_id`    | Requires a [CL ID](https://www.ebi.ac.uk/ols4/ontologies/cl).                                                                                                                                                         |
| `disease_ontology_id`      | Requires a [MONDO ID](https://www.ebi.ac.uk/ols4/ontologies/mondo).                                                                                                                                                   |
| `perturbation_ontology_id` | Must be a valid Ensembl gene ID (e.g., `ENSG00000156127`), [ChEBI ID](https://www.ebi.ac.uk/chebi/) (e.g., `CHEBI:16681`), [ChEMBL ID](https://www.ebi.ac.uk/chembl/) (e.g., `CHEMBL1234567`), or [NCBI Taxonomy ID](https://www.ncbi.nlm.nih.gov/taxonomy) (e.g., `9606`). |
| `tissue_ontology_id`       | Requires a [UBERON ID](https://www.ebi.ac.uk/ols4/ontologies/uberon).                                                                                                                                                 |


To lookup ontology terms, we recommend using the [EMBL-EBI Ontology Lookup Service](https://www.ebi.ac.uk/ols4/).

Models have a limited acceptable range of metadata input values. 
If you provide a value that is not in the acceptable range, the API will return an error.

### Making Predictions

Once your query is ready, you can send it to the API to generate gene expression data.

```{r predict-2, eval=FALSE}
# Request counts data (not log-CPM)
result <- predict_query(query, as_counts = TRUE)
```

If you want the full API response beyond just than just the result of the metadata and expression returned put `raw_response = TRUE`.

### Working with Results

```{r analyze, eval=FALSE}
# Access metadata and expression matrices
metadata <- result$metadata
expression <- result$expression

# Check dimensions
dim(expression)

# View metadata sample
head(metadata)
```

You may want to process the data in chunks or save it for later use:

```{r large-data, eval=FALSE}
# Save results to RDS file
saveRDS(result, "synthesize_results.rds")

# Load previously saved results
result <- readRDS("synthesize_results.rds")

# Export as CSV
write.csv(result$expression, "expression_matrix.csv")
write.csv(result$metadata, "sample_metadata.csv")
```


### Custom Validation

You can validate your queries before sending them to the API:

```{r validation}
# Validate structure
validate_query(query)

# Validate modality
validate_modality(query)
```

## Session info

```{r session-info}
sessionInfo()
```

## Additional Resources

- [Package Source Code](https://github.com/synthesizebio/rsynthbio)
- [File Bug Reports](https://github.com/synthesizebio/rsynthbio/issues)
